---
title: "Computer Vision with Sara"
author: "Odd Kiva"
date: "`r Sys.Date()`"
lang: en
site: "bookdown::bookdown_site"
documentclass: book
bibliography: ["phd-utf8.bib"]
biblio-style: apalike
link-citations: yes
---


# Preface {-}

This book aims at sharing my current understanding of various aspects in
computer vision during my personal learning journey as I have been maintaining
and extending the library *Sara*. The audience I am targeting here are students,
scientists or software developers who want to understand those kind of details.

Recently, I have started compiling various notes about the different
mathematical and practical aspects related to this fascinating discipline. By
snowball effect, little by little, these notes have been becoming more and more
structured as I keep learning more and more and retouching the writing here and
there.

I believe these notes carry some value and it would be a real shame to reduce
these as either mere blog posts or as parts of documentation regarding *Sara*'s
API. So this book is my modest contribution as to giving back. Hell yeah! I
would be very happy to know if it proved to be useful to even just one person.

One particular choice I made in writing this book is to depart from the dry and
impersonal style commonly used in scientific writings and use a more personal
narrative through my learning journey. As a humble computer vision practitioner
continuously learning and inevitably making mistakes along the way, this is more
natural to me. On top of that, an impersonal style would just sound boring and
not particularly engaging to younger or less experimented readers that are eager
to learn. In this book, I aim at detailing more the mathematical contents
without sacrificing the experimental aspect of computer vision. Too often,
publications would understandably gloss over some mathematical or practical
details to sell more the novelty or performance of their approaches, I have
decided to take the time to detail these for algorithms and methods that I
understand well.

> "Perfection is the enemy of good." - Voltaire.

*Finally, these notes reflect the best of my understanding as of this writing.
While I work hard to check that my contents are factually correct, I, as a human
being, am not certainly not immune to misunderstanding or errors. Should you
find any, __please pretty please__ report back to me and explain to me.*

```{r, echo = FALSE}
include_svg = function(path) {
  if (knitr::is_latex_output()) {
    output = xfun::with_ext(path, 'pdf')
    # you can compare the timestamp of pdf against svg to avoid conversion if necessary
    system2('rsvg-convert', c('-f', 'pdf', '-a', '-o', shQuote(c(output, path))))
  } else {
    output = path
  }
  knitr::include_graphics(output)
}
```

```{r, echo = FALSE}
cc_by_image_url <- 'https://i.creativecommons.org/l/by/4.0/88x31.png'
cc_by_image_file <- 'figures/cc_by_image.png'

cc_by_shield_url <- "https://img.shields.io/badge/License-CC%20BY%204.0-lightgrey.svg"
cc_by_shield_file <- "figures/cc_by_shield.svg"

if (knitr::is_latex_output() & !file.exists(cc_by_image_file)) {
  download.file(cc_by_image_url, cc_by_image_file, mode = 'wb')
} else if (!knitr::is_latex_output()) {
  cc_by_image_file <- cc_by_image_url
}

if (knitr::is_latex_output() & !file.exists(cc_by_shield_file)) {
  download.file(cc_by_shield_url, cc_by_shield_file, mode = 'wb')
} else if (!knitr::is_latex_output()) {
  cc_by_shield_file <- cc_by_image_url
}
```
The online version of this book is licensed under the [Creative Commons Attribution 4.0 International License][cc-by].

[![CC BY 4.0][cc-by-image]][cc-by]

[cc-by]: http://creativecommons.org/licenses/by/4.0/
[cc-by-image]: `r cc_by_image_file`
[cc-by-shield]: `r include_svg(cc_by_shield_file)`

## What is Sara? {-}

*Sara* is mainly a small and easy-to-use C++ computer vision library. It does
have various stubs that provide binding into other higher-level and elegant
languages like *Python* and particularly *Swift*, which strikes a good balance
between elegance and performance to me.

*Sara* (सार) is a *Sanskrit* word meaning *essence*. On a less serious note, I was
looking for a nice simple *human* name and the Spanish transliteration into this
lovely female name was then chosen intentionally.

Sara focuses on:

1. having an **easy-to-use and simple API**,
2. having **human-readable and efficient** implementations of computer vision
   algorithms,
3. **rigorous testing**.

Sara is licensed with the _[Mozilla Public License version
2.0](https://gitlab.com/DO-CV/sara/-/blob/master/LICENSE)_.

The book chooses to emphasize more on the mathematical details in the
algorithmic implementations. I believe that my API is most of the time
straightforward enough (hopefully!).

For the reference API documentation, refer to the **Doxygen**-based
documentation hosted at Codedocs.xyz: https://codedocs.xyz/DO-CV/sara/.

```{r, echo = FALSE}
codedocs_url <- 'https://codedocs.xyz/DO-CV/sara.svg'
if (!file.exists(codedocs_file <- 'figures/sara.svg'))
  download.file(codedocs_url, codedocs_file, mode = 'wb')
```

[`r include_svg(if (identical(knitr:::pandoc_to(), 'html')) codedocs_url else codedocs_file)`](https://codedocs.xyz/DO-CV/sara/)
