# Learning Journey in R

The real motivation that made me want to learn about R is because of the R
*bookdown* package [@xie:2020:bookdown], [@xie:2021:bookdown]. Compared to
Sphinx documentation, This package is stunning in terms of:

1. pretty default yet customizable aesthetics (*gitbook* !),
2. smooth and natural LaTeX support,
3. and most of all *multilanguage capabilities*, particularly C++^[C++ support is
   done via Rcpp [@eddelbuettel:2018:rcpp]. While this is not a panacea, it is
   more than sufficient for demonstration purposes in this book. In particular,
   *Eigen* [@eigenweb] is supported by the existing package *RcppEigen*
   [@eddelbuettel:2013:rcppeigen]] and Python to support and check my writings
   with *correct* and *testable* code.

So much that it pushed me to port my initial Sphinx documentation into this
technology without further ado.

With feelings of both excitement and frustration as a newbie, it took me a few
days to experiment and frantically port the documentation by finding the useful
information bits gleaned here and there in StackOverflow, GitHub and elsewhere
instead of the original documentation.

Namely *bookdown* implements *literate programming*, a term coined by
[@knuth:1984:lp]. Thus writing becomes a truly engaging for me as writing a book
actually becomes akin to a software development project where I can architect
the outline of the book and writing a section feels like writing a software
library.

Shortly after I investigated if there are any Python alternative and just learnt
about *jupyter-book* (https://jupyterbook.org) is also available. However it is
not yet clear to me whether it has multilanguage support as good as in
*bookdown*. So I will stick with *bookdown* for now.

## Cheatsheet for Code Embedding

If it is not done yet, I would recommend you to read the section 3.1 of the
online book first: https://bookdown.org/yihui/rmarkdown-cookbook/r-code.html

This section is mostly a quick-and-dirty memento for me where I list toy code
examples I found online and which I tested myself. They are copy-pasted here so
that I do not have to google them again. You can copy-paste them if you wish to
experiment on your own.

### R Example

Just like a Python notebook, R markdown allows us to write code and break it
down into multiple chunks and evaluate them as you are writing the report.
Shamelessly copying from:
https://bookdown.org/yihui/rmarkdown-cookbook/r-code.html.

```{r}
# Open this code with:
# ```{r}

x <- 5  # radius of a circle

# Close this with:
# ```
```

The R markdown file we are writing has a state, so the variable `x` written
above can be reused and evaluated in later parts of our file.

So for a circle with radius `r x`, its area is `r pi * x^2`.

### Bash Example

```{bash}
echo "Hello Bash!"
```

### C++ Example

This is the language I care the most about since it is my primary language.

- Make sure that some packages are installed, without trying to be exhaustive,
  since I am currently learning myself. Specifically on the R console:

  ```{r, eval = FALSE}
  install.packages("Rcpp")
  install.packages("RcppEigen")
  ```

- Create your C++ compile environment by creating the file `~/.R/Makevars`.

  ```
  CXXFLAGS=-std-c++17 -I/path/to/my/favorite/libraries  # etc.
  ```
  Here I just want to make sure that C++ 17 is activated.

Let us see the following example in action.

```{Rcpp, cache=TRUE}

#include <iostream>
#include <utility>
#include <string>

#include <Rcpp.h>

// [[Rcpp::plugins(cpp17)]

// [[Rcpp::export]]
int fibonacci(int x)
{
  const auto [a, b] = std::make_pair(0, 1);
  if (x == 0 || x == 1)
    return x;
  return fibonacci(x-1) + fibonacci(x-2);
}

// [[Rcpp::export]]
Rcpp::String hello_me()
{
  return Rcpp::String("hello");
}
```

```{r}
fibonacci(10L)
hello_me()
```

One caveat worth noting is that as documented in the web Rcpp does not support
*trailing return types* which I very much prefer using because of its
aesthetics.

```{Rcpp, eval = FALSE}
// Rcpp is not able to compile this code :(

// [[Rcpp::export]]
auto square(int x) -> int
{
  return x * x;
}
```

### Python Example

I really care about *Matplotlib* and *NumPy*. So here is one code example I
found online and tested myself.

```{python, cache=TRUE}
from matplotlib import pyplot as plt
from matplotlib.ticker import MultipleLocator
import numpy as np

# Two example plots
fig = plt.figure()
ax1 = fig.add_subplot(2,2,1)
ax2 = fig.add_subplot(2,2,2)

spacing = 0.5 # This can be your user specified spacing.
minorLocator = MultipleLocator(spacing)
ax1.plot(9 * np.random.rand(10))
# Set minor tick locations.
ax1.yaxis.set_minor_locator(minorLocator)
ax1.xaxis.set_minor_locator(minorLocator)
# Set grid to use minor tick locations.
ax1.grid(which = 'minor')

spacing = 1
minorLocator = MultipleLocator(spacing)
ax2.plot(9 * np.random.rand(10))
# Set minor tick locations.
ax2.yaxis.set_minor_locator(minorLocator)
ax2.xaxis.set_minor_locator(minorLocator)
# Set grid to use minor tick locations.
ax2.grid(which = 'minor')
plt.show()
```

## Visual Communication

I am mostly interested in making drawing to better convey my messages in my
writings. And one question came into my mind: can we do things quickly and
elegantly instead of *Matplotlib* [@hunter:2007:matplotlib]? After all, we are
in the realm of R.

Let us start with a mini project: make additional drawings to improve the
section about SIFT (section \@ref(sift)).

### Mini-Project: Drawings for the SIFT Section {#making-sift-drawings}

There is a lot of mathematics in this section \@ref(sift) that deserve be
supported with more drawings to better convey my messages. For one thing, I
write quite a bit about the geometry of overlapping patches with formula and
only provides formula but no drawings. So if I were a reader of the book, I
would probably have a "meh" reaction and not try to understand them at all.

#### Starting Point

I found the following code example at
https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/rect.html.

Since I am writing a book here, we can just serve the book with the following
command in the R console:

```{r, eval = FALSE}
bookdown::serve_book()
```

Let us copy-paste the code into an existing `Rmd` file and see it running in
action:

```{r, cache=TRUE, fig.align='center'}
require(grDevices)
## set up the plot region:
op <- par(bg = "thistle")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "",
     main = "2 x 11 rectangles; 'rect(100+i,300+i,  150+i,380+i)'")
i <- 4*(0:10)
## draw rectangles with bottom left (100, 300)+i
## and top right (150, 380)+i
rect(100+i, 300+i, 150+i, 380+i, col = rainbow(11, start = 0.7, end = 0.1))
rect(240-i, 320+i, 250-i, 410+i, col = heat.colors(11), lwd = i/5)
## Background alternating  ( transparent / "bg" ) :
j <- 10*(0:5)
rect(125+j, 360+j,   141+j, 405+j/2, col = c(NA,0),
     border = "gold", lwd = 2)
rect(125+j, 296+j/2, 141+j, 331+j/5, col = c(NA,"midnightblue"))
mtext("+  2 x 6 rect(*, col = c(NA,0)) and  col = c(NA,\"m..blue\")")

## an example showing colouring and shading
plot(c(100, 200), c(300, 450), type= "n", xlab = "", ylab = "")
rect(100, 300, 125, 350) # transparent
rect(100, 400, 125, 450, col = "green", border = "blue") # coloured
rect(115, 375, 150, 425, col = par("bg"), border = "transparent")
rect(150, 300, 175, 350, density = 10, border = "red")
rect(150, 400, 175, 450, density = 30, col = "blue",
     angle = -30, border = "transparent")

legend(180, 450, legend = 1:4, fill = c(NA, "green", par("fg"), "blue"),
       density = c(NA, NA, 10, 30), angle = c(NA, NA, 30, -30))

par(op)
```

#### Breaking Down the Example

Great! this code is working but there is too much information for somebody who
currently knows nothing about R and the aesthetics is awful.

Let's take it slow and simplify the code to the bare minimum, with a few things
that we experiment.

- https://www.r-bloggers.com/2017/10/loading-r-packages-library-or-require/
  recommends `library(.)` instead of `require(.)` as it makes the program easier
  to debug. So be it:

  ```{r}
  library(grDevices)
  ```

- We want to render the region defined by its bounding box. The parameter
  `type=n` apparently makes sure that the canvas is blank. The remaining
  parameters are legends.

  ```{r, cache=TRUE, fig.align='center'}
  # Let's try another background color.
  op <- par(bg = "#f7f7f7")

  # The region to render.
  x_min_max <- c(100, 250)
  y_min_max <- c(300, 450)
  plot(x_min_max, y_min_max,
       type = "n", xlab = "", ylab = "",
       main = "Main Title: One Rectangle")

  # I just want one rectangle.
  color <- "#c6c2e2"
  line_width <- 2
  rect(120, 340, 180, 410, col = color, lwd = line_width)

  # Subtitle
  mtext("Subtitle: One Rectangle")
  ```

#### Let's Go!

From now on, let's go all out to draw the geometry of overlapping patches while
referring back to the code example as a reference.

First, I want to communicate the patch overlap by means of transparent
rectangles. According to
https://stackoverflow.com/questions/18056146/how-to-add-a-transparent-rectangle-to-a-r-boxplot-plot#,
the alpha blending will do the job:
```
rect(x0, y0, x1, y1, col = rgb(0, 0, 1, alpha = 0.2))
```

To make our work manageable and easily debuggable, we will decouple the geometry
data and the drawing.

##### Geometry Data

SIFT divides the image patch into a grid $N \times N$ of smaller patches. Let
us define the sequence of indices.

```{r}
N <- 4
j <- 0:(N - 1)
i <- 0:(N - 1)
```

The patch center is:
```{r}
patch_center <- c(0, 0)
```

Let us consider the following point in a the image patch to illustrate our point
in Section \@ref(sift).

```{r}
point_in_patch <- c(-0.8, 1.1)
```

Each subpatch $(i, j)$ has their center at the following grid coordinates:
```{r}
xi <- i - (N - 1) / 2
yj <- j - (N - 1) / 2
```

```{r, echo = FALSE}
xi
yj
```

The patch region is exactly
```{r}
region_min_max_x <- c(-(N-1)/2 - 1, (N-1)/2 + 1)
region_min_max_y <- c(-(N-1)/2 - 1, (N-1)/2 + 1)
```

```{r, echo = FALSE}
region_min_max_x
region_min_max_y
```

#### Geometry Rendering
Moving onto the rendering, let's make the viewing domain slightly larger than
the patch region:

```{r}
x_view_min_max <- c(region_min_max_x[1] - 1, region_min_max_x[2] + 1)
y_view_min_max <- c(region_min_max_y[1] - 1, region_min_max_y[2] + 1)
```

```{r, echo = FALSE}
x_view_min_max
y_view_min_max
```

##### Communicating the Pitfalls Due to the Lack of Overlap

We are now rendering the geometry of patches when they do not overlap each other
with the following code.

```{r, cache=TRUE, fig.align='center'}
# Define an empty canvas.
op <- par(bg = "#f7f7f7", pty="s")
plot(x_view_min_max, y_view_min_max,
     type = "n",
     xlab = "x", ylab = "y",
     main = "SIFT Grid Geometry")

# The radius of each subpatch (i, j).
r <- 0.5

# Cosmetics.
rect_line_width <- 0.5
rect_color <- rgb(0.8, 0.8, 0.8, alpha = 0.5)

for (x in xi) {
  # Render in batch.
  rect(x - r, yj - r, x + r, yj + r,
       col = rect_color,
       lwd = rect_line_width)
}

# Image patch center.
points(patch_center[1], patch_center[2], col="red")

# Patch centers.
for (x in xi) {
  for (y in yj)
    points(x, y, col = "blue")
}

# Draw the 4 closest subpatch to the point in the image patch.
rect(xi[1] - r, yj[4] - r, xi[1] + r, yj[4] + r,
     col = rgb(1, 0, 0, alpha = 0.2),
     lwd = 1)
rect(xi[2] - r, yj[4] - r, xi[2] + r, yj[4] + r,
     col = rgb(0, 1, 0, alpha = 0.2),
     lwd = 1)
rect(xi[1] - r, yj[3] - r, xi[1] + r, yj[3] + r,
     col = rgb(0, 0, 1, alpha = 0.2),
     lwd = 1)
rect(xi[2] - r, yj[3] - r, xi[2] + r, yj[3] + r,
     col = rgb(0, 1, 1, alpha = 0.2),
     lwd = 1)

points(point_in_patch[1], point_in_patch[2],
       col="red",  # In red color please
       pch="+",    # I want a "+" shape
       cex=2)      # Thicker size please


mtext("Without Overlapping")
```

So far we have drawn the $N \times N$ grid that divides the patch and crossed in
red a particular point in the patch, close to the border of the green patch. We
will focus on the legend later.

##### Animation

It would be great if we could animate the drawing would communicate better my
point if the patches were not overlapping. This is possible according to:
https://bookdown.org/yihui/rmarkdown-cookbook/animation.html#ref-R-gifski

In this case, let us use `polygon` instead of `rect` to generalise the drawing.
First we want to refactor the rendering code with the little R knowledge at our
hand.

```{r, cache=TRUE}
draw_patch <- function(center, theta, scale, subpatch_radius,
                       color = "red") {
  # Alias the variable.
  r <- subpatch_radius

  # Cosmetics.
  rect_color <- rgb(0.8, 0.8, 0.8, alpha = 0.5)
  rect_line_width <- 0.5

  # Transformation matrix
  T <- scale * matrix(c(cos(theta), -sin(theta), center[1],
                        sin(theta),  cos(theta), center[2],
                                 0,           0,         1),
                      nrow = 3,
                      byrow = TRUE)

  # Render the grid of patches.
  for (x in xi) {
    for (y in yj) {
      # Enumerate the quad vertices.
      X <- matrix(c(x - r, x + r, x + r, x - r,
                    y - r, y - r, y + r, y + r,
                        1,     1,     1,     1),
                  nrow = 3,
                  byrow = TRUE)

      # Transform the vertices.
      TX <- T %*% X

      # Separate the coordinates.
      u <- TX[1,]
      v <- TX[2,]
      polygon(u, v, col = rect_color, lwd = rect_line_width)
    }
  }

  # Draw the center of the image patch.
  points(center[1], center[2], col = color)

  # Draw the center of each subpatch (i, j).
  for (x in xi) {
    for (y in yj) {
      Txy <- T %*% matrix(c(x, y, 1), nrow=3)
      u <- Txy[1]
      v <- Txy[2]
      points(u, v, col = color)
    }
  }

  # Draw the closest subpatches (i, j) to the point in the patch.
  for (k in 1:2) {
    for (l in 3:4) {
      # Enumerate the quad vertices.
      X <- matrix(c(xi[k] - r, xi[k] + r, xi[k] + r, xi[k] - r,
                    yj[l] - r, yj[l] - r, yj[l] + r, yj[l] + r,
                            1,         1,         1,         1),
                  nrow = 3,
                  byrow = TRUE)

      # Transform the vertices.
      TX <- T %*% X
      u <- TX[1,]
      v <- TX[2,]

      # Draw the polygon.
      if (k == 1 & l == 3)
        color <- rgb(1, 0, 0, alpha = 0.2)
      if (k == 2 & l == 3)
        color <- rgb(0, 1, 0, alpha = 0.2)
      if (k == 1 & l == 4)
        color <- rgb(0, 0, 1, alpha = 0.2)
      if (k == 2 & l == 4)
        color <- rgb(0, 1, 1, alpha = 0.2)
      polygon(u, v, col = color, lwd = 1)
    }
  }

  # Draw the point in the patch.
  points(point_in_patch[1], point_in_patch[2],
         col = "darkblue",  # In another color please
         pch = "+",              # I want a "+" shape
         cex = 2)                # Thicker size please
}
```

Finally, after a bit of debugging, we can reuse the function at will and
generate an animation that illustrates the border effect when the sub-patches do
not overlap and when the detector detects the keypoint in a noisy manner

```{r, animation.hook="gifski", interval=0.1, cache=TRUE, fig.align='center'}
op <- par(bg = "#f7f7f7", pty="s")

for (ii in 1:30) {
  # Reset the canvas
  plot(x_view_min_max, y_view_min_max,
       type = "n",                      # Reset the canvas
       xlab = "x", ylab = "y",          # Legend
       main = "SIFT Grid Geometry")     # Title.

  patch_center_shifted <- c(patch_center[1] + shift_x[ii],
                            patch_center[2] + shift_y[ii])

  # The ground truth image patch.
  draw_patch(patch_center, 0, 1, 0.5,
             color = "slateblue")

  # The noisy image patch.
  draw_patch(patch_center_shifted, theta[ii], scale_inc[ii], 0.5,
             color = "orangered")

  mtext("Without Overlapping")
}

for (ii in 1:30) {
  # Reset the canvas
  plot(x_view_min_max, y_view_min_max,
       type = "n",                      # Reset the canvas
       xlab = "x", ylab = "y",          # Legend
       main = "SIFT Grid Geometry")     # Title.

  patch_center_shifted <- c(patch_center[1] + shift_x[ii],
                            patch_center[2] - 0.5 * shift_y[ii])

  # The ground truth image patch.
  draw_patch(patch_center, 0, 1, 0.5,
             color = "slateblue")

  # The noisy image patch.
  draw_patch(patch_center_shifted, -theta[ii], scale_dec[ii], 0.5,
             color = "orangered")

  mtext("Without Overlapping")
}
```

##### Communicating about the Overlapping

We have worked hard to create the useful rendering function and can profit from
it to illustrate the benefit of overlapping patches.

Let us animate the dilation of the patch radius to illustrate that sub-patches
now overlap each other.

```{r, animation.hook="gifski", interval=0.1, cache=TRUE, fig.align='center'}
op <- par(bg = "#f7f7f7", pty="s")

patch_radius <- c(seq(0.5, 0.9, length.out = 5),
                  seq(0.9, 0.99, length.out = 5),
                  seq(0.99, 1, length.out = 20))

for (ii in 1:30) {
  # Reset the canvas
  plot(x_view_min_max, y_view_min_max,
       type = "n",                      # Reset the canvas
       xlab = "x", ylab = "y",          # Legend
       main = "SIFT Grid Geometry")     # Title.

  # The ground truth image patch.
  draw_patch(patch_center, 0, 1, patch_radius[ii],
             color = "slateblue")

  mtext("Increase of the Patch Radius for the Patch Overlapping")
}
```

Now we can communicate the robustness we benefit due to the overlapping between
sub-patches with another animation.

We are going to vary the following parameters.

```{r}
theta <- seq(0, 0.2, length.out = 30)
scale_inc <- seq(1, 1.3, length.out = 30)
scale_dec <- seq(1, 0.8, length.out = 30)
shift_x <- seq(0, 0.5, length.out = 30)
shift_y <- seq(0, 0.3, length.out = 30)
```

We can now generate the animation in GIF format.

```{r, animation.hook="gifski", interval=0.1, cache=TRUE, fig.align='center'}
op <- par(bg = "#f7f7f7", pty="s")

subpatch_radius <- 1
theta <- seq(0, 0.2, length.out = 30)
scale <- seq(1, 1.3, length.out = 30)
shift_x <- seq(0, 0.5, length.out = 30)
shift_y <- seq(0, 0.3, length.out = 30)

for (ii in 1:30) {
  # Reset the canvas
  plot(x_view_min_max, y_view_min_max,
       type = "n",                      # Reset the canvas
       xlab = "x", ylab = "y",          # Legend
       main = "SIFT Grid Geometry")     # Title.

  patch_center_shifted <- c(patch_center[1] + shift_x[ii],
                            patch_center[2] + shift_y[ii])

  # The ground truth image patch.
  draw_patch(patch_center, 0, 1,
             subpatch_radius,
             color = "slateblue")

  # The noisy image patch.
  draw_patch(patch_center_shifted, theta[ii], scale_inc[ii],
             subpatch_radius,
             color = "orangered")

  mtext("With Overlapping")
}

for (ii in 1:30) {
  # Reset the canvas
  plot(x_view_min_max, y_view_min_max,
       type = "n",                      # Reset the canvas
       xlab = "x", ylab = "y",          # Legend
       main = "SIFT Grid Geometry")     # Title.

  patch_center_shifted <- c(patch_center[1] + shift_x[ii],
                            patch_center[2] - 0.5 * shift_y[ii])

  # The ground truth image patch.
  draw_patch(patch_center, 0, 1,
             subpatch_radius,
             color = "slateblue")

  # The noisy image patch.
  draw_patch(patch_center_shifted, -theta[ii], scale_dec[ii],
             subpatch_radius,
             color = "orangered")

  mtext("With Overlapping")
}
```

##### Conclusion

We have completed the most technical part of this mini-project, which is to
illustrate the different parameters that conditions SIFT's anatomy through
various animations. This mini-project turns out to be quite time-consuming yet
rewarding and most of all, it has highlighted that visual communication largely
remains an art.

Much remains to be improved and could be done later while I am learning on the
way.

## Linear Algebra Cheatsheet

Array broadcasting is wonderful in NumPy:

```{python}
import numpy as np

a = np.arange(0, 10).reshape((2, 5))
b = np.arange(0, 2).reshape((2, 1))

a + b
```

In R we have to use a more verbose way to achieve the equivalent operation using
`sweep`

```{r}
a <- matrix(0:9, nrow = 2, byrow = TRUE)
b <- matrix(0:1, nrow = 2)

sweep(a, 1, b, "+")
```

A bit disappointing in terms in conciseness but it works.
