# Blobs

```{r}
#| echo: false
#| warning: false
#| label: sift-on-cheetah
#| out.width: 80%
#| fig.align: center
#| fig.cap: >
#|   Black dots on the cheetah's fur are one beautiful pattern among
#|   infinitely many in Nature.
#|   Sara's SIFT implementation retrieves these particularly well as most of
#|   them can be detected as dark circular blobs in this photograph.
#|   Credits: https://www.britannica.com/animal/cheetah-mammal.
knitr::include_graphics(paste0(getwd(), "/features/cheetah-2.jpg"))
```

```{r}
#| echo: false
#| warning: false
#| label: sift-on-sunflower-field
#| out.width: 80%
#| fig.align: center
#| fig.cap: >
#|   The SIFT detector implemented in Sara responds particularly well on the
#|   sunflower heads in this image depicting a sunflower field. They are
#|   detected as dark circular blobs surrounded by a brighter background. Notice
#|   how with the scale-space representation of the image, we are able to
#|   estimate well the size of the circular flower head.
knitr::include_graphics(paste0(getwd(), "/features/sunflower-field.jpg"))
```

In this chapter, we will review the geometry of local features that *SIFT*
[@Lowe:2004:ijcv], *SURF* [@BayETVG:2008:cviu], MSER [@MatasCUP:2002:bmvc] and
affine-covariant regions [@MikolajczykTSZMSKVG:2005:ijcv] share in common.

Figures \@ref(fig:sift-on-cheetah) and \@ref(fig:sift-on-sunflower-field) are
iconic illustrations of SIFT blobs.

These features were very utilized for high-level object recognition tasks before
the widespread adoption of deep learning and are still playing a key part in
structure-from-motion pipelines that I feel compelled to say a few words about
them.

N.B.: this chapter contains bits of my thesis [@Ok:2013:phdthesis], which I have
simplified.

## Geometry

In general any locally invariant regions, *a.k.a.* features, present a common
set of geometric characteristics, be it SIFT, SURF, MSER, or Harris-Affine etc.
We will detail these in the sequel. Now in multiple view geometry and
structure-from-motion, they tend to be viewed as mere points in the images but
they are actually much more than that.

Before moving forward in this chapter, let us denote a feature by the italic
text letter $x$ for convenience.

More than just being a point indeed, a feature $x$ is fully characterized by:

- a center with pixel coordinates $\x \in \mathbb{R}^2$,
  which in practice is the only information that structure-from-motion pipelines
  use,

- a shape $\Shape_x$, which represents the (possibly anisotropic) scale of
  feature $x$ and can be summarized^[MSER is a special case as I'll explain
  further below] as an ellipse as follows:

  \begin{equation}
    \Shape_x \eqdef \left\{ \x' \in \mathbb{R}^2\ \left|\ (\x'-\x)^T \bSigma_x
  (\x'-\x) \leq 1 \right. \right\},
  \end{equation}

  where $\bSigma_x \in \mathbb{R}^{2\times 2}$ is a scale matrix.

- a dominant gradient orientation $\theta_x$ if it is calculated.

While this may seem redundant with the shape definition, we can also derive the
following definition^[I doubt you will see this definition anywhere else, so
this is my small innovation.].

:::puzzle
```{definition, name="Scale-Normalized Signed Distance Function"}
The scale matrix $\bSigma_x$ also allows to define a *scale-normalized signed
distance function* to $x$:

\begin{equation}
  d_x(\x') \eqdef (\x' - \x)^T \bSigma (\x' - \x) - 1,
\end{equation}

where the zero level set is the *elliptic* shape of the feature.
```
:::

Let us illustrate some kinds of feature that have been used in previous research
to set things clear in our mind.

### Locally Similar-Invariant Features

An example of a SIFT feature is illustrated in Figure \@ref(fig:sift-feature).
To be more accurate, what SIFT detects are *local extrema of Difference of
Gaussians*, which we term simply as DoG for brevity. By design, a DoG feature is
circular and only guarantees invariance to similarity transforms
(scale+rotation) in contrast to other affine invariant features presented in
[@MikolajczykTSZMSKVG:2005:ijcv].

```{r}
#| echo: false
#| warning: false
#| label: sift-feature
#| out.width: 80%
#| fig.align: center
#| fig.cap: >
#|   Geometry of a SIFT feature.
#|   Credits: http://www.vlfeat.org/overview/sift.html.
knitr::include_graphics(paste0(getwd(), "/features/feature.jpg"))
```

Other similarity-invariant detectors enjoy the same similarity invariance as
SIFT. Let us cite for example SURF [@BayETVG:2008:cviu], Edge Foci
[@ZitnickR:2011:iccv], SFOP [@ForstnerDS:2009:iccv]. I don't know much about
them but I really like the elegant geometric properties that SFOP and Edge Foci
enjoy.^[And I should study them someday.]

### Locally Affine-Invariant Features

In contrast to DoG-SIFT keypoints, the shape of affine-covariant features is
elliptic. Such an elliptic shape is obtained by an algorithmic process called
affine shape adaptation, which iteratively refines the localization of the
initial feature detection. We will not go into details about this method and
will be content with illustrating such features with Figure
\@ref(fig:affine-covariant-regions).

```{r}
#| echo: false
#| warning: false
#| label: affine-covariant-regions
#| out.width: 80%
#| fig.align: center
#| fig.cap: >
#|   Locally Affine-Invariant Features.
#|   Credits: [@MikolajczykTSZMSKVG:2005:ijcv]
knitr::include_graphics(paste0(getwd(), "/features/affine-covariant-regions.jpg"))
```

As shown in Figure \@ref(fig:affine-covariant-regions), it is quite compelling
that most of the time the affine shape is well estimated across different change
of viewpoints and would bring a superior advantage in the feature matching.

In my experience however, I found that SIFT still performs very well without the
affine shape adaptation [@MikolajczykTSZMSKVG:2005:ijcv] and actually better
than affine-invariant features in structure-from-motion. In fact, we can even
tell that by closely inspecting the features in Figure
\@ref(fig:affine-covariant-regions): some Harris-Affine or Hessian-Affine
features seem to have their center and shape really off by a few noticeable
pixels between the two viewpoints in this figure.^[Optical illusion? I don't
think so... my fear is that these are far from being uncommon to the point of
harming the 3D reconstruction, but this claim needs proper investigation.].

One thing to note about MSER elliptic shapes: they are not obtained from affine
shape adaptation but from the computation of the second-moment matrix.
Regardless of their elliptic shape, MSERs are not very popular in
structure-from-motion pipelines mostly because they tend to be too few at the
detection.

Besides the affine shape adaptation has a significant computational burden and
is tricky to implement, which leads us to think this might not be worth going
through the pain.


### MSER: A Special Case.

In fact, MSER features [@MatasCUP:2002:bmvc] are more than just an elliptic
shape: they are a set of connected pixels. That allows us to represent a much
richer set of shapes, and we can use them for letter recognition. See for
example [@GomezK:2014:icpr].

As I carefully said it above, a MSER feature is *summarized* more compactly as
an elliptic region by calculating a second-order moment matrix. To do so, we
cumulate the locations of all the pixels $\mathbf{x}'$ in the MSER as follows:

\begin{equation}
  \bSigma_x^{-1} = \frac{1}{\mathrm{area}(x)}
  \displaystyle \int (\x' - \x)(\x' - \x)^T \mathop{d}\x'
\end{equation}

The reason for which I use the integral $\int$ was to avoid the confusion with
the discrete sum symbol $\sum$ with the scale symbol $\bSigma$.

Implementation-wise, denoting the $x$- and $y$-coordinates of each 2D vector by
$\x_u$ and $\x_v$, this is simply the following discrete sums:

\begin{equation}
  \bSigma_x^{-1} = \frac{1}{\mathrm{area}(x)}
  \begin{bmatrix}
    \displaystyle \sum_{\mathbf{x}'} (\mathbf{x}'_u - \mathbf{x}_u)^2 &
    %
    \displaystyle \sum_{\mathbf{x}'}
      (\mathbf{x}'_u - \mathbf{x}_u)(\mathbf{x}'_v - \mathbf{x}_v) \\
    %
    \displaystyle \sum_{\mathbf{x}'}
      (\mathbf{x}'_u - \mathbf{x}_u)(\mathbf{x}'_v - \mathbf{x}_v) &
    %
    \displaystyle \sum_{\mathbf{x}'} (\mathbf{x}'_v - \mathbf{x}_v)^2
  \end{bmatrix},
\end{equation}

the second-order moment matrix being also the inverse of the scale matrix.

### Personal Conclusion

I conclude this section with a personal (biased) note since I am a SIFT fan.

Despite the advances in feature detection, SIFT remains a very good option.
Besides the SIFT patent has expired, which now makes it viable for commercial
applications. The default parameters that comes with it are so good and well
validated over the widespread adoption in research. If parameter tuning becomes
necessary, it is very easy to do so as I found them quite straightforward to
interpret.

Although only similarity-invariant, SIFT still performs very well compared to
affine-covariant regions in my experience. My main complaint with
affine-covariant features is the messiness I have to deal after the detection.
Usually, I have to filter them with non-maxima suppression to remove clusters of
features.^[Sorry for being too critical... but yes I was a long-time user of
affine-covariant features.]

That is not the case with SIFT, MSER where the keypoints are very distinctive.
